---
title: "Exploratory Analysis"
author: "Azim Abdul Wahid, Hannah Chung, Michelle To, Jamie Costales"
date: "5/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("../scripts/enrollment.R")
source("../scripts/exemptions.R")
source("../scripts/counties.R")
source("../scripts/districts_by_county.R")

library(ggplot2)
library(shiny)
library(knitr)
library(plotly)
```

## Introduction

```{r intro, echo =FALSE }

```

## Summary Information

```{r summary, echo = FALSE}

```

## Enrollment and Vaccines

```{r enrollment, echo=FALSE}
enrollment_perc_complete_plotly
```

## Reasons for No Vaccination

```{r exemptions, echo=FALSE}
exemption_vis
```

We wanted to know more about the distribution of the reasons for not vaccinating. We thought that if we knew the distribution, we would know where to best focus efforts in addressing the lack of vaccination.

This pie chart displays the count and percentage of the different reasons for not vaccinating. We can deduce that not vaccinating due to personal exemption makes up 71.6% (44,104 children) of the total number of children (61,598 children) who didn’t get vaccinated. We see that we should be focusing our efforts in targeting children who have personal exemptions.

## Immunization by County

```{r counties, echo=FALSE}
counties_plot
```

## Immunization for Districts by County

```{r districts_by_county, echo=FALSE, comment=NA, results='hide'}
districts_plot
```

